---
title: Blog 2
date: 'September 16, 2024'
slug: prediction-2
---

```{r, include = FALSE, message = FALSE, warning = FALSE}
# load packages and wrangle data
library(blogdown)
library(tidyverse)
library(car)
library(tidyr)
library(lubridate)
library(readr)

theme_nick <- theme_bw() + 
  theme(panel.border = element_blank(),
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        plot.title = element_text(size = 15, hjust = 0.5), 
        legend.position = "bottom",
        legend.text = element_text(size = 10))

## set working directory here
setwd("~/Documents/GOV 1347/GOV 1347/content/post/2024-09-11-prediction-2")

####----------------------------------------------------------#
#### Read, merge, and process data.
####----------------------------------------------------------#

# Load popular vote data. 
d_popvote <- read_csv("popvote_1948-2020.csv")

# Load economic data from FRED: https://fred.stlouisfed.org. 
# Variables, units, & ranges: 
# GDP, billions $, 1947-2024
# GDP_growth_quarterly, %
# RDPI, $, 1959-2024
# RDPI_growth_quarterly, %
# CPI, $ index, 1947-2024
# unemployment, %, 1948-2024
# sp500_, $, 1927-2024 
d_fred <- read_csv("fred_econ.csv")

# Load economic data from the BEA: https://apps.bea.gov/iTable/?reqid=19&step=2&isuri=1&categories=survey#eyJhcHBpZCI6MTksInN0ZXBzIjpbMSwyLDMsM10sImRhdGEiOltbImNhdGVnb3JpZXMiLCJTdXJ2ZXkiXSxbIk5JUEFfVGFibGVfTGlzdCIsIjI2NCJdLFsiRmlyc3RfWWVhciIsIjE5NDciXSxbIkxhc3RfWWVhciIsIjIwMjQiXSxbIlNjYWxlIiwiMCJdLFsiU2VyaWVzIiwiUSJdXX0=.
# GDP, 1947-2024 (all)
# GNP
# RDPI
# Personal consumption expenditures
# Goods
# Durable goods
# Nondurable goods
# Services 
# Population (midperiod, thousands)
d_bea <- read_csv("bea_econ.csv") |> 
  rename(year = "Year",
         quarter = "Quarter", 
         gdp = "Gross domestic product", 
         gnp = "Gross national product", 
         dpi = "Disposable personal income", 
         consumption = "Personal consumption expenditures", 
         goods = "Goods", 
         durables = "Durable goods", 
         nondurables = "Nondurable goods", 
         services = "Services", 
         pop = "Population (midperiod, thousands)")

# Filter and merge data. 
d_inc_econ <- d_popvote |> 
  filter(incumbent_party == TRUE) |> 
  select(year, pv, pv2p, winner) |> 
  left_join(d_fred |> filter(quarter == 2)) |> 
  left_join(d_bea |> filter(quarter == "Q2") |> select(year, dpi))

```

**September 16, 2024** - Today marks 50 days until the 2024 Presidential Election. For this week's model updates, I am focusing on one of the fundamentals of political forecasting: the economy. One of the most famous quotes in the political world belongs to longtime Democratic strategist James Carville, who famously said "It's the economy, stupid", and that statement is abundantly true. Americans evaluate the economy and choose candidates based on the recent state of the economy, known as voting retrospectively (Achen and Bartels 2017). Historically, economic metrics from Quarter 2 leading up to the election are strongly correlated with incumbent party vote share, as illustrated by the figures below. The blue line depicts the line of best fit for all years included in the data, white the red line represents the line of best fit for all elections included besides 2020, which was a major outlier on many economic indicators. In general, we can see that incumbent presidents, or their party successors, are rewarded if there was recent growth in Gross Domestic Product. 


```{r, echo = FALSE, warning = FALSE, message = FALSE}
# show correlation between economic factors and incumbent vote share
d_inc_econ_no2020 <- d_inc_econ %>%
  filter(year != 2020)

p1 <- d_inc_econ |> 
  ggplot(aes(x = GDP_growth_quarterly, y = pv2p, label = year)) + 
  geom_hline(yintercept = 50, lty = 2) + 
  geom_vline(xintercept = 0.01, lty = 2) +
  labs(x = "Second Quarter GDP Growth (%)", 
       y = "Incumbent Party's National Popular Vote Share") + 
  geom_smooth(method = "lm", se = FALSE) + 
  geom_smooth(data = d_inc_econ_no2020, aes(x= GDP_growth_quarterly, y = pv2p), method = "lm", color = "red", se = FALSE) + 
  geom_text() + 
  theme_bw()

p2 <- d_inc_econ |> 
  ggplot(aes(x = RDPI_growth_quarterly, y = pv2p, label = year)) + 
  geom_hline(yintercept = 50, lty = 2) + 
  geom_vline(xintercept = 0.01, lty = 2) +
  labs(x = "Second Quarter RDPI Growth (%)", 
       y = "Incumbent Party's National Popular Vote Share") + 
  geom_smooth(method = "lm", se = FALSE) + 
  geom_smooth(data = d_inc_econ_no2020, aes(x= RDPI_growth_quarterly, y = pv2p), method = "lm", color = "red", se = FALSE) + 
  geom_text() + 
  theme_bw()

p3 <- d_inc_econ |> 
  ggplot(aes(x = CPI, y = pv2p, label = year)) + 
  geom_hline(yintercept = 50, lty = 2) + 
  geom_vline(xintercept = 0.01, lty = 2) +
  labs(x = "Consumer Price Index", 
       y = "Incumbent Party's National Popular Vote Share") + 
  geom_smooth(method = "lm", se = FALSE) + 
  geom_smooth(data = d_inc_econ_no2020, aes(x= CPI, y = pv2p), method = "lm", color = "red", se = FALSE) + 
  geom_text() + 
  theme_bw()

p1
```
In my model forecast for this week, I assess how key economic indicators influence peoples’ votes; constructing an OLS linear regression model using Quarter 2 (April, May, June) economic data from 1976 to 2020 as a proxy for perceptions of the economy leading up to the election to predict Democratic candidate two party vote share in each state. My national economic data includes GDP growth, real disposable personal income growth (RDPI), Consumer Price Index (CPI) to cover three highly salient and different aspects of evaluating the economy; national macroeconomic growth, personal finance situation, and inflation, respectively. However, because of the size of the United States, national metrics may not capture individual state trends. Therefore, my model includes average monthly unemployment for the three months of Q2 for each state to try to capture the more localized economic conditions. 

In addition to these economic variables, I built upon my model last week by including a lagged variable representing Democrat two-party vote share in the previous presidential election, to account for the influence of partisanship in addition to economic conditions. This is why I chose Democrat two-party vote share to simply predict the winner of each state, as it would be more complicated to create a lagged incumbent party vote share variable instead of just choosing one of the two parties. To account my choice of focusing on one party’s variables and incumbency bias, I include two indicator variables: ‘inc_prez’ that denotes if the current president is running for reelection, and ‘inc_party’ to denote the party of the incumbent president, as voters may hold the incumbent president or political party responsible for current economic conditions. 

```{r, include = FALSE, message = FALSE, warning = FALSE}
# build model to predict states

statewide_pv <- read_csv("clean_wide_state_2pv_1948_2020.csv")

unemployment_state <- read_csv("unemployment_by_state_NEW.csv")

```

```{r, include = FALSE}
# cleaning unemployment data
unemployment_state <- unemployment_state %>% 
  mutate(DATE = as.Date(DATE, format = "%Y-%m-%d")) %>%
  mutate(year = year(DATE),
         month = month(DATE),
         day = day(DATE)) %>%
  pivot_longer(cols = starts_with("ALUR") | starts_with("AKUR") | starts_with("AZUR") | starts_with("ARUR") | 
                     starts_with("CAUR") | starts_with("COUR") | starts_with("CTUR") | starts_with("DEUR") | 
                     starts_with("FLUR") | starts_with("GAUR") | starts_with("HIUR") | starts_with("IDUR") | 
                     starts_with("ILUR") | starts_with("INUR") | starts_with("IAUR") | starts_with("KSUR") | 
                     starts_with("KYUR") | starts_with("LAUR") | starts_with("MEUR") | starts_with("MDUR") | 
                     starts_with("MAUR") | starts_with("MIUR") | starts_with("MNUR") | starts_with("MSUR") | 
                     starts_with("MOUR") | starts_with("MTUR") | starts_with("NEUR") | starts_with("NVUR") | 
                     starts_with("NHUR") | starts_with("NJUR") | starts_with("NMUR") | starts_with("NYUR") | 
                     starts_with("NCUR") | starts_with("NDUR") | starts_with("OHUR") | starts_with("OKUR") | 
                     starts_with("ORUR") | starts_with("PAUR") | starts_with("RIUR") | starts_with("SCUR") | 
                     starts_with("SDUR") | starts_with("TNUR") | starts_with("TXUR") | starts_with("UTUR") | 
                     starts_with("VTUR") | starts_with("VAUR") | starts_with("WAUR") | starts_with("WVUR") | 
                     starts_with("WIUR") | starts_with("WYUR") | starts_with("DCUR"),
               names_to = "state_code", values_to = "unemployment_rate") %>%
  mutate(state = case_when(
    state_code == "ALUR" ~ "Alabama",
    state_code == "AKUR" ~ "Alaska",
    state_code == "AZUR" ~ "Arizona",
    state_code == "ARUR" ~ "Arkansas",
    state_code == "CAUR" ~ "California",
    state_code == "COUR" ~ "Colorado",
    state_code == "CTUR" ~ "Connecticut",
    state_code == "DEUR" ~ "Delaware",
    state_code == "FLUR" ~ "Florida",
    state_code == "GAUR" ~ "Georgia",
    state_code == "HIUR" ~ "Hawaii",
    state_code == "IDUR" ~ "Idaho",
    state_code == "ILUR" ~ "Illinois",
    state_code == "INUR" ~ "Indiana",
    state_code == "IAUR" ~ "Iowa",
    state_code == "KSUR" ~ "Kansas",
    state_code == "KYUR" ~ "Kentucky",
    state_code == "LAUR" ~ "Louisiana",
    state_code == "MEUR" ~ "Maine",
    state_code == "MDUR" ~ "Maryland",
    state_code == "MAUR" ~ "Massachusetts",
    state_code == "MIUR" ~ "Michigan",
    state_code == "MNUR" ~ "Minnesota",
    state_code == "MSUR" ~ "Mississippi",
    state_code == "MOUR" ~ "Missouri",
    state_code == "MTUR" ~ "Montana",
    state_code == "NEUR" ~ "Nebraska",
    state_code == "NVUR" ~ "Nevada",
    state_code == "NHUR" ~ "New Hampshire",
    state_code == "NJUR" ~ "New Jersey",
    state_code == "NMUR" ~ "New Mexico",
    state_code == "NYUR" ~ "New York",
    state_code == "NCUR" ~ "North Carolina",
    state_code == "NDUR" ~ "North Dakota",
    state_code == "OHUR" ~ "Ohio",
    state_code == "OKUR" ~ "Oklahoma",
    state_code == "ORUR" ~ "Oregon",
    state_code == "PAUR" ~ "Pennsylvania",
    state_code == "RIUR" ~ "Rhode Island",
    state_code == "SCUR" ~ "South Carolina",
    state_code == "SDUR" ~ "South Dakota",
    state_code == "TNUR" ~ "Tennessee",
    state_code == "TXUR" ~ "Texas",
    state_code == "UTUR" ~ "Utah",
    state_code == "VTUR" ~ "Vermont",
    state_code == "VAUR" ~ "Virginia",
    state_code == "WAUR" ~ "Washington",
    state_code == "WVUR" ~ "West Virginia",
    state_code == "WIUR" ~ "Wisconsin",
    state_code == "WYUR" ~ "Wyoming",
    state_code == "DCUR" ~ "District Of Columbia")) %>%
  select(year, month, day, state, unemployment_rate)
```

```{r, include = FALSE}
unemployment_state <- unemployment_state %>%
  filter(month >= 4 & month <= 6) %>%
  filter(year %in% c(1976, 1980, 1984, 1988, 1992, 1996, 2000, 2004, 2008, 2012, 2016, 2020, 2024)) %>%
  mutate(quarter = 2) %>%
  group_by(year, quarter, state) %>%
  summarize(ur_avg = mean(unemployment_rate))

econ_ur_join <- unemployment_state %>%
  left_join(d_inc_econ, by = c("year", "quarter")) %>%
  rename(pv_national = pv,
         pv2p_national = pv2p,
         winner_national = winner)

full_data_statewide <- econ_ur_join %>%
  left_join(statewide_pv, by = c("year", "state")) %>%
  mutate(inc_prez = ifelse(year %in% c(1976, 1980, 1984, 1996, 2004, 2012, 2020), 1, 0)) %>%
  mutate(inc_party = ifelse(year %in% c(1980, 1996, 2000, 2012, 2016, 2024), "D", "R"))

```


```{r, include = FALSE}
train <- full_data_statewide %>% filter(year != 2024)
d_bea <- d_bea %>% mutate(year = as.character(year), quarter = as.character(quarter))
d_fred <- d_fred %>% mutate(year = as.character(year), quarter = as.character(quarter))

# fix lags for 2024
only_2020 <- statewide_pv %>% filter(year == 2020) %>%
  select(state, D_pv2p, D_pv2p_lag1) %>%
  mutate(lag_D_pv2p_2020 = D_pv2p, lag_D_pv2p_2016 = D_pv2p_lag1)

test <- full_data_statewide %>% filter(year == 2024) %>%
  mutate(year = as.character(year), quarter = as.character(quarter)) %>%
  left_join(only_2020, by = "state") %>%
  left_join(d_bea, by = c("year", "quarter")) %>%
  left_join(d_fred, by = c("year", "quarter")) %>%
  mutate(GDP_growth_quarterly = GDP_growth_quarterly.y, RDPI_growth_quarterly = RDPI_growth_quarterly.y, CPI = CPI.y, D_pv2p_lag2 = lag_D_pv2p_2016, D_pv2p_lag1 = lag_D_pv2p_2020)
  


mod1 <- lm(D_pv2p ~ GDP_growth_quarterly + RDPI_growth_quarterly + CPI + ur_avg + inc_party + inc_prez + D_pv2p_lag1, data = train)

mod1

```

```{r, include = FALSE, message = FALSE, warning = FALSE}
test$output_D_pv2p <- predict(mod1, newdata = test)

test <- test %>%
  mutate(state_winner = ifelse(output_D_pv2p > 50, "D", "R"))
```

```{r, include = FALSE, message = FALSE, warning = FALSE}
# predict electoral college with map

ec <- read_csv("corrected_ec_1948_2024.csv") %>%
  mutate(year = as.character(year))

test <- test %>% 
  left_join(ec, by = c("state", "year"))

test_outcome <-  test %>%
  group_by(state_winner) %>%
  summarize(electoral_votes = sum(electors))
```
```{r, include = FALSE}
# create electoral map
library(maps)
states_map <- map_data("state")

test$state <- tolower(test$state)
results_map <- merge(states_map, test, by.x = "region", by.y = "state", all.x = TRUE)


```

Applying my individual state model including these variables to 2024 data yields dramatically different results than last week’s simple model. My new model, including national economic indicators, state-level unemployment, previous results, and incumbency factors predicts that Republican former President Donald Trump wins the 2024 Presidential Election with **326** electoral votes compared to Democrat Vice President Kamala Harris’ **212** electoral votes. In other words, voters are generally forecasted to punish the incumbent party, the Democratic Party, for the current economic conditions, as Kamala Harris is the current vice president to Democratic President Joe Biden. However, many crucial swing states had predicted Democratic two-party vote shares just below the 50% mark, which means that a Democratic victory is within the margin of error.

```{r, echo = FALSE}
ggplot(data = results_map, aes(x = long, y = lat, group = group)) +
  geom_polygon(aes(fill = output_D_pv2p), color = "black") +  
  coord_map("polyconic") + 
scale_fill_gradient2(low = "red", mid = "white", high = "blue", midpoint = 50, 
                       breaks = c(20, 35, 50, 65, 80), 
                       limits = c(25, 75),
                     name = "Democratic Two-Party Vote Share") + theme_void() + theme(legend.position = "bottom") +
  labs(title = "Model Forecast of 2024 Presidential Election* - 9/16/24",
       caption = "*Alaska and Hawaii not included")
```

While this may seem like a lopsided forecast, a thousand repetition cross-validation results in a relatively small mean out-of-sample error of -1.129, which indicates that this model has a good fit on this data. Plotting the residuals and a QQ-plot of all the data demonstrate that this model fits the data fairly appropriately, as residuals look decently evenly distributed. 


```{r, echo = FALSE}
res <- resid(mod1)
plot(fitted(mod1), res)
abline(0,0)

qqnorm(res)
qqline(res)
```

```{r, include = FALSE}
#evaluate model
set.seed(2024)
out_samp_errors <- sapply(1:1000, function(i) {
  years_out_samp <- sample(train$year, 1) 
  mod1_test <- lm(D_pv2p ~ GDP_growth_quarterly + RDPI_growth_quarterly + CPI + ur_avg + inc_party + inc_prez + D_pv2p_lag1, train[!(train$year %in% years_out_samp),])
  out_samp_pred <- predict(mod1_test, train[train$year %in% years_out_samp,])
  out_samp_truth <- train$D_pv2p[train$year %in% years_out_samp]
  mean(out_samp_pred - out_samp_truth)
})

mean(out_samp_errors)
```

Additionally, I chose state-level unemployment to capture state economic conditions because I had limited data options for this week’s assignment. In my data there is an extremely weak correlation between state unemployment and vote share ( *r* = 0.13). Perhaps in future models I may find different local economic indicators to better capture these state level effects. 


```{r, include = FALSE}
cor(full_data_statewide$ur_avg, full_data_statewide$D_pv2p, use = "complete.obs")
```

```{r, echo = FALSE}
full_data_statewide |> 
  ggplot(aes(x = ur_avg, y = D_pv2p)) + 
  geom_hline(yintercept = 50, lty = 2) + 
  geom_vline(xintercept = 0.01, lty = 2) +
  labs(x = "State Unemployment Rate (%)", 
       y = "Democrat Statewide Two-Party Voteshare",
       title = "Statewide Unemployment and Voteshare (1976-2020)") + 
  geom_smooth(method = "lm", se = FALSE) + 
  geom_point() + 
  theme_bw()
```
